Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-14 text/manuscript-full-KS.qmd
went through introduction

---
title: "Structuring College Access: The Market Segment Model and College Board Geomarkets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

[^1]: On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.

@@

Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary for their progeny to claim spots at selective colleges [@RN1228; @RN1241; @RN4991] [ADD CITES]. @RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [@RN861] \[cite others\]. Scholarship on enrollment management offers an agentic depiction of organizational behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324].

We argue that these perspectives ignore an important mechansim of social reproduction: third-party vendors create products that sort students on behalf of college. These products often amplify historic inequality in educational opportunity by using past snapshots of student demand to make recommendations about where colleges should prioritize recruiting efforts. In the contemporary era of private equity funded platform capitalism, third-party recruiting products have become increasingly powerful, granular, and ubiquitous. However, we argue that a watershed moment in their development was the creation of Geomarkets.

@@

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board underwrote the project -- providing data and funding -- and published the book as part of their nascent efforts "to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. Conceptualized as an effort to to "capture and quantify" (p. 11) the knowledge of admissions officers, @RN4982 developed the Market Segment Model by analyzing the SAT score-sending behavior of high school seniors in 1980. The thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets, a territory of an admissions recruiter. For instance, Figure X panel A shows Geomarkets in the Chicago-land area. As a planning tool for colleges, the Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

@@

Geomarkets have been incorporated into the supply-side structure of college access in three ways. Anecdotally, Geomarkets became an organizing principle for how college admissions offices allocate admissions recruiters to territories. Figure X panel B shows that Chicago Geomarkets define recruiting territories for University of Chicago admissions recruiters (Appendix Figures BLAH BLAH). Second, Geomarkets are the basis for the College Board Enrollment Planning Service (EPS). EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets.

@@

Third, Geomarkets were incorporated into College Board's student list product, Student Search Service. Student lists contain the contact information of prospective students and have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling names [@belkin2019-studata; @list_biz]. @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include or exclude prospects from particular Geomarkets.

@@

Scholarship on enrollment management has focused on college behavior [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. However, scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising given sociology is concerned with social sorting [@RN4873], particularly sorting into college.

@@

Our analyses address the two research questions, which speak to how Geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between Geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining Geomarket shapefiles to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular Geomarkets? This research question is motivated by @RN4982, which recommends that selective colleges target affluent Geomarkets. Additionally, prior research on enrollment management behavior finds that selective colleges focus recruiting visits on high schools in affluent, predominantly white communities [@RN3519; @RN4758]. We analyze this research question using data from actual student list purchases, which were collected using public records requests. We analyze student list purchases that include all Geomarkets in a metropolitan area in order to simulate which prospects would be included or excluded had the student list purchase filtered on particular Geomarkets.

The following section introduces background and scholarship about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. \[We find that\ldots ?PARAGRAPH ON RESULTS?\]

Finally, we discuss implications for scholarship and for policy. The sociology of education should move beyond the fixation on schools and policies as the primary sorting machines. @RN527 \[p. 148\] remind us that the organizational field includes the totality of relevant actors." Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [CITE]. By investigating high-leverage actors and products in the organizational field, sociology can develop important insights that shape future federal and state regulatory policy. \[REVISE THIS PARAGRAPH\]

@@

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The Student Search Service -- created in 1972 and copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from a student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

Second, technological advances creates new means of identifying and serving prospects. College Board began selling the contact information of test-takers in 1972 [@belkin2019-studata]. ACT followed suit and the two testing companies enjoyed a near duopoly for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded news sources of student list data. Another data source is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than used by more than 10 million K-12 students and by 40% ofUS high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based targeting.

>>>>

---
title: "Structuring College Access: The Market Segment Model and College Board Geomarkets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)", "Crystal Han (Stanford University)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."^[On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

@@

Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary to claim spots at selective colleges [e.g., @RN1228; @RN1241; @RN4991]. <!--@RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. -->On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [e.g., @RN865; @RN861]. Scholarship on enrollment management offers an agentic depiction of college behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324].

We argue that these perspectives ignore an important mechansim of social reproduction: third-party vendors create products that sort students on behalf of college. These products often amplify historic inequality in educational opportunity by using historical snapshots of student demand to make recommendations about where colleges should prioritize recruiting efforts. In the contemporary era of private equity funded platform capitalism, third-party recruiting products have become increasingly powerful, granular, and ubiquitous. However, a watershed moment in their development was the creation of College Board Geomarkets.

@@

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board underwrote the project -- providing data and funding -- and published the book as part of their nascent efforts "to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. Conceptualized as an effort to to "capture and quantify" (p. 11) the knowledge of admissions officers, @RN4982 developed the Market Segment Model by analyzing the SAT score-sending behavior of high school seniors in 1980. The thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets. Figure X panel A shows Geomarkets in the Chicago-land area. As a planning tool for colleges, the Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

@@

Geomarkets have been incorporated into the supply-side structure of college access in three ways. Anecdotally, Geomarkets became an organizing principle for how college admissions offices allocate admissions recruiters to territories. For example, Figure X panel B shows that Chicago Geomarkets define recruiting territories for University of Chicago admissions recruiters. Second, Geomarkets are the basis for the College Board Enrollment Planning Service (EPS). EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets.

@@

Third, Geomarkets were incorporated into College Board's student list product, Student Search Service. Student lists contain the contact information of prospective students and have been the primary source of "lead generation" in U.S. higher education since 1972, when College Board began selling names [@belkin2019-studata; @list_biz; @RN4728]. <!-- @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. --> Colleges control which prospect profiles they purchase by selecting search filters, such as SAT score, AP score, state, etc. Geomarket filters enable colleges to include or exclude prospects from particular Geomarkets.

@@

Scholarship on enrollment management has focused on the behavior of colleges [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. However, scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products, their consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with social sorting [@RN4873], particularly sorting into college.

@@

Our analyses address the two research questions, which speak to how Geomarkets are utilized within EPS software (RQ1) and within the Student Search Service student list product (RQ2). First, what is the socioeconomic and racial variation between Geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining Geomarket shapefiles to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus excluded prospects vary when student list purchases filter on particular Geomarkets? This research question is motivated by @RN4982, which recommend that selective colleges target affluent Geomarkets, and by prior research on enrollment management behavior, which finds that selective colleges focus recruiting visits on high schools in affluent, predominantly white communities [@RN3519; @RN4758]. We analyze this research question using data from actual student list purchases, which we collected using public records requests. We analyze student list purchases that include all Geomarkets in a metropolitan area in order to investigate which prospects would be included or excluded had the student list purchase filtered on particular Geomarkets.

The following section introduces background and scholarship about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. \[We find that\ldots ONE OR TWO SENTENCES\]. Finally, we discuss implications for scholarship. Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [vista_eab_ml]. We propose two fruitful approaches to studying the role of third-party vendors in enrollment management. First, case studies -- like this manuscript and @RN5039 -- utilize primary data collection to conduct granular analyses of particular products or vendors. Second, a growing number of subscription databases (e.g., Orbis, Pitchbook, LSEG) serve the information needs of the investment community by providing data on private firms and acquisitions or investments in private firms. These data sources may provide the basis for a scalable "knowledge infrastructure" [@RN5025] that investigates the role of third-party vendors in education.

@@

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The Student Search Service -- copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from a student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

Second, technological advances creates new means of identifying and serving prospects. College Board and ACT leveraged data on test-takers to enjoy a near duopoly in market for student list data, one that lasted for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded news sources of student list data. Another data source is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than used by more than 10 million K-12 students and by 40% ofUS high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based targeting.
